### 2013-12-23 article.tex
Template for a PeerJ paper

\documentclass[12pt]{article}
\usepackage{graphicx, subfigure, float, amsmath, amssymb, color}
\usepackage[margin = 1.0 in]{geometry}
\usepackage{natbib}

\title{Accessing biological data with semantic web technologies}
\author{Egon L.\ Willighagen$^1$}
\begin{document}

\date{\today}
\maketitle

\noindent
$^1$ Department of Bioinformatics - BiGCaT, NUTRIM, Maastricht University, Maastricht, The Netherlands

\bigskip
\noindent
$^*$Corresponding author\\
$\phantom{^*}$Email: egon.willighagen@maastrichtuniversity.nl\\
$\phantom{^*}$Twitter: @egonwillighagen\\
$\phantom{^*}$ORCID: http://orcid.org/0000-0001-7542-0286

\bigskip
\noindent
Manuscript type: research article

\bigskip
\noindent Keywords: semantic web, workflow

\newpage
\begin{abstract}
\end{abstract}

\section{Introduction}
\label{Introduction}

Foo~\cite{Willighagen2011}.

\bibliographystyle{peerj} % from: https://github.com/clauswilke/protein_design_and_site_variability/blob/master/latex/peerj.bst
\bibliography{article}

\end{document}

### 2013-12-23 article.tex
Initial intro, moved to biblatex, and some refs

@@
 \usepackage[margin = 1.0 in]{geometry}
-\usepackage{natbib}
+\usepackage[backend=biber,firstinits=true,sorting=none]{biblatex}
+\addbibresource{article.bib}
 
 \title{Accessing biological data with semantic web technologies}
-\author{Egon L.\ Willighagen$^1$}
+\author{Egon L.\ Willighagen$^1,2$}
 \begin{document}
@@
 $^1$ Department of Bioinformatics - BiGCaT, NUTRIM, Maastricht University, Maastricht, The Netherlands
+$^2$ Department of Molecular Toxicology, IMM, Karolinska Institutet, Stockholm, Sweden
 
@@
 \section{Introduction}
-\label{Introduction}
 
-Foo~\cite{Willighagen2011}.
+Semantic Web technologies are finding their way into biological
+databases~\cite{Belleau2008,Chen2010,Samwald2011,Williams2012,Willighagen2013}. The recent launch of the
+Resource Description Framework (RDF) services at the European Bioinformatics Institute is a good example
+of the impact it has on the field, despite the adoption taking place for longer already~\cite{EBIRDFPlatform}.
+At the same time, most new databases do not yet use semantic web technologies, which may be partly
+caused by the lack of availability of tools to handle RDF data.
 
-\bibliographystyle{peerj} % from: https://github.com/clauswilke/protein_design_and_site_variability/blob/master/latex/peerj.bst
-\bibliography{article}
+RDF allows the combination of a machine readable framework with the use of ontologies that allow
+sharing data and the meaning in the some data exchange format. This is enabled by a set of open
+specifications, such as the underlying RDF, RDF Schema, the Web Ontology Language, and serialization formats
+such as RDF/XML, N-TRIPLES, and Turtle. Crucial, however, seems the specification of the SPARQL
+query language that allow extracting data from local or remote RDF triple stores~\cite{Seaborne2009}. In fact,
+federated SPARQL queries allow extracting data from multiple triple stores at the same time.
+
+At the same time, R has established itself as a key tool for statistics, particularly in the biological
+sciences. Many packages have been developed for this domain and are available from CRAN and Bioconductor,
+allowing to combine various bioinformatics and data analysis approaches. These packages include,
+for example, biomaRt, which have the purpose of extracting information from a remote database.
+
+We here introduce two packages, rrdflibs and rrdf, for the statistics environment R, based on the Apache
+Jena library and the RJava interface to Java~\cite{RMan,Jena,RJava}. Using this package we show how
+data aggregation can be performed from remote SPARQL end points, and how local data in other formats
+can be converted into RDF to querying with SPARQL.
+
+\section{Materials \& Methods}
+
+The rrdf packages provides RDF and SPARQL functionality by exposing that functionality from the
+Apache Jena library~\cite{Jena}. The rrdflibs package contains the Apache Jena libraries and
+nothing more. This package is close to 7~MiB large but does not change frequently. The rrdf
+package contains the R functions that wrap around the Jena functionality and convert data structures
+where needed. This package is about 112~kiB but changes more frequently. Using this approach,
+the Comprehensive R Archive Network (CRAN) server has the least amount of download throughput.
+The Jena functionality is accessible using the rJava library~\cite{RJava} which handles loading
+of the Java archives and provides methods to instantiate classes and call methods.
+
+The source code of the package is available from GitHub at \url{http://github.com/egonw/rrdf/} and
+binary packages are available from CRAN at \url{http://cran.r-project.org/web/packages/rrdf/}.
+The first patch is from March 2011 while the most recent patches are from late 2013 when the
+rrdflibs package was updated for Apache Jena 2.11.
+
+\section{Results and Discussion}
+
+The rrdf package provides basic and less basic functionality for aggregation and analysis of RDF data.
+The package can be install and loaded with these commands:
+
+\begin{verbatim}
+install.packages(c("rrdf", "rrdflibs"))
+library(rrdf)
+\end{verbatim}
+
+\subsection{Triple stores}
+
+\subsection{SPARQL}
+
+\section{Conclusions}
+
+\section{Acknowledgements}
+
+The people who have provided feedback on the rrdf package are kindly thanked for their efforts in
+getting into contact with me.
+
+\printbibliography
 

### 2013-12-24 article.tex
More example code

@@
 specifications, such as the underlying RDF, RDF Schema, the Web Ontology Language, and serialization formats
-such as RDF/XML, N-TRIPLES, and Turtle. Crucial, however, seems the specification of the SPARQL
+such as RDF/XML, N-TRIPLES, Notation3 (N3), and Turtle. Crucial, however, seems the specification of the SPARQL
 query language that allow extracting data from local or remote RDF triple stores~\cite{Seaborne2009}. In fact,
@@
 
+To handle triples, we first need a triple store. At this moment only in-memory stores are
+supported, though Jena's TDB also provides on-disk triple stores; this puts limitations to
+the amount of data you can analyze in one study.
+The package supports two kinds of stores, one that has minimal ontology support, and one
+basically just handles triples. Both can be created with the \textit{new.rdf} command:
+
+\begin{verbatim}
+ontStore = new.rdf()
+store = new.rdf(ontology=FALSE)
+\end{verbatim}
+
+When data is not read from a file or downloaded, but created from data in another format,
+such as an R matrix, individual triples can be added to a store. This works for both
+object properties and data properties. The former option links a subject to an object
+resource, while the latter links a subject to a literal value:
+
+\begin{verbatim}
+add.triple(store,
+  subject="http://example.org/Subject",
+  predicate="http://example.org/Predicate",
+  object="http://example.org/Object"
+)
+add.data.triple(store,
+  subject="http://example.org/Subject",
+  predicate="http://example.org/Predicate",
+  data="Literal value"
+)
+\end{verbatim}
+
+Data can be loaded from file by providing a file name and a format ("RDF/XML", "TURTLE",
+"N-TRIPLES", and "N3"):
+
+\begin{verbatim}
+store = load.rdf("file.n3", format="N3")
+\end{verbatim}
+
+\subsection{Example}
+
 \subsection{SPARQL}
 
+Local (in-memory) triple stores can be queried with the \emph{sparql.rdf} method by providing
+it with the store to query and a string with the SPARQL query. For example, to get all resource
+types, we can use this:
+
+\begin{verbatim}
+sparql.rdf(store,
+  paste(
+    "SELECT DISTINCT ?type WHERE {",
+    " [] a ?type ",
+    "}"
+  )
+) 
+\end{verbatim}
+
+Remote SPARQL end point can be queried in a similar fashion: we replace the triple store with the
+URL pointing to the SPARQL end point. For example, if we want to
+extract all properties of the CHEMBL615603 assay from ChEMBL~\cite{OveringtonXXXX,Willighagen2013}
+we can use the following remote query:
+
+\begin{verbatim}
+results = sparql.remote(
+  "http://rdf.farmbio.uu.se/chembl/sparql",
+  paste(
+    "SELECT DISTINCT ?predicate ?object WHERE {",
+    " ?assay <http://www.w3.org/2000/01/rdf-schema#label> \"CHEMBL615603\" ;",
+    "        ?predicate ?object .",
+    "}"
+  )
+) 
+\end{verbatim}
+
+The results object is a matrix with the variable names from the SPARQL query as column names.
+This allows us to get a predicates with:
+
+\begin{verbatim}
+results[,"predicate"]
+\end{verbatim}
+
+\subsection{Example}
+
+...
+
 \section{Conclusions}
+
+The rrdflibs and rrdf packages bring semantic web technologies to R statistical environment.
+While only a subset of Apache Jena is currently exposed, it provides key methods to deal with
+RDF data and resources.
 

### 2013-12-26 article.tex
Working out the first example: SNP data from a BioMart

@@
 We here introduce two packages, rrdflibs and rrdf, for the statistics environment R, based on the Apache
-Jena library and the RJava interface to Java~\cite{RMan,Jena,RJava}. Using this package we show how
+Jena library and the RJava interface to Java~\cite{RMan,Jena,RJava}. Compared to the existing SPARQL
+package, this package adds general RDF data handling~\cite{VanHage2013}. Using this package we show how
 data aggregation can be performed from remote SPARQL end points, and how local data in other formats
@@
 
+As an example, we will here create RDF for a small data set with information on five single nucleotide
+polymorphisms (SNPs) in the BRCA1 gene, extracted with biomaRt~\cite{Durinck2005}:
+
+\begin{verbatim}
+library(biomaRt)
+mart = biomaNoy2009Rt::useMart(biomart="snp", dataset="hsapiens_snp")
+brca1 = c("rs16940","rs16941", "rs16942", "rs799916", "rs799917")
+attribs = c(
+  "refsnp_id", "chr_name", "chrom_start",
+  "validated", "pmid_20137", "allele"
+)
+data = biomaRt::getBM(
+  attributes=attribs, filters=c("snp_filter"), values=brca1, mart=mart
+)
+\end{verbatim}
+
+Following the guidelines for generating RDF outlined by Open PHACTS~\cite{Haupt2013,Williams2012}, we identify concepts
+and matching ontology terms, using BioPortal~\cite{Noy2009}: SNP, gene, chromosome,
+article, and allele. Additionally, there are the following properties: a SNP identifier, a PubMed identifier,
+a chromosome number, a chromosomal sequence position, and a list of validations (e.g. HapMap, 1000Genomes).
+The latter will not be further semantically specified, but just be converted into a RDF literal as it is
+provided by the BioMart.
+
+Taking advantage of
+
 \subsection{SPARQL}
@@
 URL pointing to the SPARQL end point. For example, if we want to
-extract all properties of the CHEMBL615603 assay from ChEMBL~\cite{OveringtonXXXX,Willighagen2013}
+extract all properties of the CHEMBL615603 assay from ChEMBL~\cite{Gaulton2012,Willighagen2013}
 we can use the following remote query:
@@
 \end{verbatim}
+
+Jena parses the SPARQL query too, which requires the query to be valid
+against both Jena and the SPARQL end point. Because the SPARQL specification has tool specif
+extensions and that not all tools support the full of SPARQL 1.1, it can sometimes be tricky
+to find the mutually support SPARQL command subset. The rrdf package also allows to bypass
+Jena and query the end point directly by using \emph{sparql.remote(..., jena=FALSE)},
+removing the problem and allowing you to use extensions of the end point.
 

### 2013-12-26 article.tex
Fixed the affiliation data layout

@@
 \title{Accessing biological data with semantic web technologies}
-\author{Egon L.\ Willighagen$^1,2$}
+\author{Egon L.\ Willighagen$^{1,2}$}
 \begin{document}
@@
 \noindent
-$^1$ Department of Bioinformatics - BiGCaT, NUTRIM, Maastricht University, Maastricht, The Netherlands
+$^1$ Department of Bioinformatics - BiGCaT, NUTRIM, Maastricht University, Maastricht, The Netherlands \\
 $^2$ Department of Molecular Toxicology, IMM, Karolinska Institutet, Stockholm, Sweden
@@
 
-Taking advantage of
+Taking advantage of ...
 

### 2013-12-26 article.tex
Mapping concepts to ontologies...

@@
 
+\begin{table}
+\caption{A list of ontology terms for the concepts in the BRCA1 SNP example. Ontology acronyms and the prefixes
+used are explained in the text.}
+\begin{center}
+\begin{tabular}{l|l|l|l}
+ \textbf{concept} & \textbf{ontology} & \textbf{term} & \textbf{URI} \\
+\hline
+ SNP & SIO & snp & sio:SIO\_010027 \\
+\end{tabular}
+\end{center}
+\end{table}
+
 Taking advantage of ...
+SIO is the Semanticscience Integrated Ontology;
+... Prefixes used in the term Uniform Resource Identifiers (URIs) include sio for \url{http://semanticscience.org/resource/} ...
 

### 2013-12-26 article.tex
First code to create RDF from the BioMart data

@@
  SNP & SIO & snp & sio:SIO\_010027 \\
-\end{tabular}
+ gene & SO & gene & obo:SO\_0000704 \\
+ chromosome & GO & chromosome & obo:GO\_0005694 \\
+ article &  BIBO & article & bibo:Article \\
+ allele & SO & allele & obo:SO\_0001023 \\
+ SNP identifier & TMO & snp identifier & tmo:TMO\_0161 \\
+ PubMed identifier & CHEMINF & PubMed identifier & sio:CHEMINF\_000302\\
+ chromosome number & TMO & chromosome nuber & tmo:TMO\_0157 \\
+ chromosomal sequence position & TMO & chromosomal sequence position & tmo:TMO\_0122 \\
+\end{tabular}CiTO
 \end{center}
@@
 
-Taking advantage of ...
-SIO is the Semanticscience Integrated Ontology;
-... Prefixes used in the term Uniform Resource Identifiers (URIs) include sio for \url{http://semanticscience.org/resource/} ...
+Taking advantage of existing ontologies. The choice is somewhat arbitrary and no ontological
+analysis has been done, e.g. on specified domain and ranges of predicates and implied
+properties of classes. The following ontologies are used:
+SIO is the Semanticscience Integrated Ontology; TMO is the Translational Medicine Ontology;
+SO is the Sequence Types and Features Ontology; GO is the Gene Ontology;
+BIBO is the Bibliographic Ontology; CiTO is the Citation Typing Ontology;
+... Prefixes used in the term Uniform Resource Identifiers (URIs) include
+\textit{sio} for \url{http://semanticscience.org/resource/},
+\textit{tmo} for \url{http://www.w3.org/2001/sw/hcls/ns/transmed/},
+\textit{obo} for \url{http://purl.obolibrary.org/obo/},
+\textit{bibo} for \url{http://purl.org/ontology/bibo/},
+\textit{cito} for \url{} ...
+
+Using these ontology terms we can define the translations of the BioMart-extracted
+data into RDF:
+
+\begin{verbatim}
+snpStore = new.rdf(ontology=FALSE)
+rdfTypePred = "http://www.w3.org/1999/02/22-rdf-syntax-ns#type"
+snpClass = "http://semanticscience.org/resource/SIO_010027"
+createEntry <- function(row) {
+  snpID = row[1]
+  snpSubject = paste("http://example.org/snp/", snpID, sep="")
+  add.triple(snpStore,
+    subject=snpSubject, predicate=rdfTypePred, object=snpClass
+  )
+}
+\end{verbatim}
+
+We can then just iterate over the actual data with:
+
+\begin{verbatim}
+apply(data, MARGIN=1, FUN=createEntry(row) )
+\end{verbatim}
 
@@
 
+%\subsection{Example: BioPortal}
+%
+%In the BRCA1 SNP example, we found ontology terms for concepts using BioPortal. Now, at the
+%time of writing BioPortal has an experimental SPARQL end point which allows us to find terms
+%in this way too. For example, to find terms related to SNPs, we can run this code in R:
+%
+%\begin{verbatim}
+%findSNPterms = paste(
+%  "PREFIX rdfs: <http://www.w3.org/2000/01/rdf-schema#>",
+%  "SELECT DISTINCT ?term ?label WHERE {",
+%  " ?term rdfs:label ?label .",
+%  "FILTER (CONTAINS (LCASE(str(?label)), \"snp\"))",
+%  "}"
+%)
+%sparql.remote(
+%  endpoint="http://sparql.bioontology.org/sparql",
+%  sparql=findSNPterms, jena=FALSE
+%)
+%\end{verbatim}
+
 \section{Conclusions}

### 2013-12-30 article.tex
All SNP data is now converted to RDF

@@
 
+\begin{footnotesize}
 \begin{verbatim}
@@
 \end{verbatim}
+\end{footnotesize}
 
@@
 \end{verbatim}
+\end{footnotesize}
+
+We can also take advantage from the fact that R assigns values to method parameters based on
+the order in which they are given. Therefore, we can simplify the above code to:
+
+\begin{footnotesize}
+\begin{verbatim}
+add.triple(store,
+  "http://example.org/Subject",
+  "http://example.org/Predicate",
+  "http://example.org/Object"
+)
+add.data.triple(store,
+  "http://example.org/Subject",
+  "http://example.org/Predicate",
+  "Literal value"
+)
+\end{verbatim}
+\end{footnotesize}
 
@@
 \begin{center}
+\begin{footnotesize}
 \begin{tabular}{l|l|l|l}
@@
  SNP & SIO & snp & sio:SIO\_010027 \\
- gene & SO & gene & obo:SO\_0000704 \\
  chromosome & GO & chromosome & obo:GO\_0005694 \\
@@
  chromosomal sequence position & TMO & chromosomal sequence position & tmo:TMO\_0122 \\
-\end{tabular}CiTO
+\end{tabular}
+\end{footnotesize}
 \end{center}
@@
 properties of classes. The following ontologies are used:
-SIO is the Semanticscience Integrated Ontology; TMO is the Translational Medicine Ontology;
-SO is the Sequence Types and Features Ontology; GO is the Gene Ontology;
-BIBO is the Bibliographic Ontology; CiTO is the Citation Typing Ontology;
+SIO is the Semanticscience Integrated Ontology;
+TMO is the Translational Medicine Ontology;
+%SO is the Sequence Types and Features Ontology;
+GO is the Gene Ontology;
+BIBO is the Bibliographic Ontology;
+CiTO is the Citation Typing Ontology;
 ... Prefixes used in the term Uniform Resource Identifiers (URIs) include
@@
 \textit{bibo} for \url{http://purl.org/ontology/bibo/},
-\textit{cito} for \url{} ...
+\textit{cito} for \url{http://purl.org/spar/cito/} ...
+
+We can define these prefixes in the store with the following code,
+the first prefix in the full syntax and the other namespaces
+leaving out the parameters names:
+
+\begin{footnotesize}
+\begin{verbatim}
+snpStore = new.rdf(ontology=FALSE)
+add.prefix(snpStore,
+  prefix="sio", namespace="http://semanticscience.org/resource/"
+)
+add.prefix(snpStore, "tmo", "http://www.w3.org/2001/sw/hcls/ns/transmed/")
+add.prefix(snpStore, "obo", "http://purl.obolibrary.org/obo/")
+add.prefix(snpStore, "bibo", "http://purl.org/ontology/bibo/")
+add.prefix(snpStore, "cito", "http://purl.org/spar/cito/")
+add.prefix(snpStore, "snp", "http://example.org/snp/")
+add.prefix(snpStore, "art", "http://example.org/article/")
+add.prefix(snpStore, "loc", "http://example.org/location/")
+add.prefix(snpStore, "all", "http://example.org/allele/")
+add.prefix(snpStore, "ex", "http://example.org/onto/")
+add.prefix(snpStore, "pubmedid", "http://example.org/pubmed/")
+add.prefix(snpStore, "snpid", "http://example.org/snpid/")
+\end{verbatim}
+\end{footnotesize}
+
+Furthermore, we can define a number of classes and predicate, for reduced code:
+
+\begin{footnotesize}
+\begin{verbatim}
+snpClass = "http://semanticscience.org/resource/SIO_010027"
+chromosomeClass = "http://purl.obolibrary.org/obo/GO_0005694"
+pubmedIdClass = "http://semanticscience.org/resource/CHEMINF_000302"
+snpIdClass = "http://www.w3.org/2001/sw/hcls/ns/transmed/TMO_0161"
+articleClass = "http://purl.org/ontology/bibo/Article"
+alleleClass = "http://purl.obolibrary.org/obo/SO_0001023"
+
+rdfTypePred = "http://www.w3.org/1999/02/22-rdf-syntax-ns#type"
+isDescribedBy = "http://purl.org/spar/cito/isDescribedBy"
+onChromosome = "http://example.org/onto/onChromosome"
+hasAttribute = "http://semanticscience.org/resource/CHEMINF_000200"
+hasStart = "http://example.org/onto/hasStart"
+hasValue = "http://semanticscience.org/resource/SIO_000300"
+hasValidation = "http://example.org/onto/hasValidation"
+hasAllele = "http://example.org/onto/hasAllele"
+hasLocation = "http://example.org/onto/hasLocation"
+\end{verbatim}
+\end{footnotesize}
 
@@
 
+\begin{footnotesize}
 \begin{verbatim}
-snpStore = new.rdf(ontology=FALSE)
-rdfTypePred = "http://www.w3.org/1999/02/22-rdf-syntax-ns#type"
-snpClass = "http://semanticscience.org/resource/SIO_010027"
 createEntry <- function(row) {
   snpID = row[1]
+  chr = row[2]
+  chrStart = row[3]
+  validation = row[4]
+  pubmedID = row[5]
+  allele = row[6]
+
   snpSubject = paste("http://example.org/snp/", snpID, sep="")
-  add.triple(snpStore,
-    subject=snpSubject, predicate=rdfTypePred, object=snpClass
+  pubmedObject = paste("http://example.org/article/a", pubmedID, sep="")
+  alleleObject = paste("http://example.org/allele/", snpID, sep="")
+  chrObject = paste("http://example.org/location/", snpID, sep="")
+
+  snpIDObject = paste("http://example.org/snpid/", snpID, sep="")
+  pubmedIDObject = paste("http://example.org/pubmed/a", pubmedID, sep="")
+
+  add.triple(snpStore, snpSubject, rdfTypePred, snpClass)
+  add.triple(snpStore, snpSubject, hasAttribute, snpIDObject)
+  add.triple(snpStore, snpIDObject, rdfTypePred, snpIdClass)
+  add.data.triple(snpStore, snpIDObject, hasValue, snpID)
+  # validation information
+  sapply(unlist(strsplit(validation, split=",")),
+    function(validationItem) {
+      if (!is.null(validationItem)) {
+        add.data.triple(snpStore,
+          snpSubject, hasValidation, validationItem
         )
       }
+    }
+  )
+  # allele information
+  add.triple(snpStore, snpSubject, hasAllele, alleleObject)
+  add.triple(snpStore, alleleObject, rdfTypePred, alleleClass)
+  add.data.triple(snpStore, alleleObject, hasValue, allele)
+  # chromosome location information
+  add.triple(snpStore, snpSubject, hasLocation, chrObject)
+  add.triple(snpStore, chrObject, rdfTypePred, chromosomeClass)
+  add.data.triple(snpStore, chrObject, onChromosome, chr)
+  add.data.triple(snpStore, chrObject, hasStart, chrStart)
+  # pubmed information
+  add.triple(snpStore, snpSubject, isDescribedBy, pubmedObject)
+  add.triple(snpStore, pubmedObject, rdfTypePred, articleClass)
+  add.triple(snpStore, pubmedObject, hasAttribute, pubmedIDObject)
+  add.triple(snpStore, pubmedIDObject, rdfTypePred, pubmedIdClass)
+  add.data.triple(snpStore, pubmedIDObject, hasValue, pubmedID)
+}
 \end{verbatim}
+\end{footnotesize}
 
-We can then just iterate over the actual data with:
+We can then just iterate over the actual data with and safe it to a file:
 
+\begin{footnotesize}
 \begin{verbatim}
-apply(data, MARGIN=1, FUN=createEntry(row) )
+apply(data, MARGIN=1, FUN=createEntry)
+save.rdf(snpStore, filename="test.n3", format="N3")
 \end{verbatim}
+\end{footnotesize}
+
+Or output is as a Notation3 string with:
+
+\begin{footnotesize}
+\begin{verbatim}
+cat(asString.rdf(snpStore))
+\end{verbatim}
+\end{footnotesize}
 

### 2013-12-30 snp.R

@@ -1,0 +1,95 @@
+snpStore = new.rdf(ontology=FALSE)
+add.prefix(snpStore,
+  prefix="sio", namespace="http://semanticscience.org/resource/"
+)
+add.prefix(snpStore, "tmo", "http://www.w3.org/2001/sw/hcls/ns/transmed/")
+add.prefix(snpStore, "obo", "http://purl.obolibrary.org/obo/")
+add.prefix(snpStore, "bibo", "http://purl.org/ontology/bibo/")
+add.prefix(snpStore, "cito", "http://purl.org/spar/cito/")
+add.prefix(snpStore, "snp", "http://example.org/snp/")
+add.prefix(snpStore, "art", "http://example.org/article/")
+add.prefix(snpStore, "loc", "http://example.org/location/")
+add.prefix(snpStore, "all", "http://example.org/allele/")
+add.prefix(snpStore, "ex", "http://example.org/onto/")
+add.prefix(snpStore, "pubmedid", "http://example.org/pubmed/")
+add.prefix(snpStore, "snpid", "http://example.org/snpid/")
+
+
+
+
+snpClass = "http://semanticscience.org/resource/SIO_010027"
+chromosomeClass = "http://purl.obolibrary.org/obo/GO_0005694"
+pubmedIdClass = "http://semanticscience.org/resource/CHEMINF_000302"
+snpIdClass = "http://www.w3.org/2001/sw/hcls/ns/transmed/TMO_0161"
+articleClass = "http://purl.org/ontology/bibo/Article"
+alleleClass = "http://purl.obolibrary.org/obo/SO_0001023"
+
+rdfTypePred = "http://www.w3.org/1999/02/22-rdf-syntax-ns#type"
+isDescribedBy = "http://purl.org/spar/cito/isDescribedBy"
+onChromosome = "http://example.org/onto/onChromosome"
+hasAttribute = "http://semanticscience.org/resource/CHEMINF_000200"
+hasStart = "http://example.org/onto/hasStart"
+hasValue = "http://semanticscience.org/resource/SIO_000300"
+hasValidation = "http://example.org/onto/hasValidation"
+hasAllele = "http://example.org/onto/hasAllele"
+hasLocation = "http://example.org/onto/hasLocation"
+
+
+
+
+
+createEntry <- function(row) {
+  snpID = row[1]
+  chr = row[2]
+  chrStart = row[3]
+  validation = row[4]
+  pubmedID = row[5]
+  allele = row[6]
+
[…]

### 2013-12-30 article.tex
A working remote SPARQL example

@@
 
-Or output is as a Notation3 string with:
+Or output it as a Notation3 string with:
 
@@
 
-...
+Remote SPARQLing can be used to get more information about biological entities of interest. The following
+example retreives information from the Uniprot database using the experimental SPARQL end point. In
+particular, it retrieves annotations for proteins encoded by the BRCA1 gene:
 
-%\subsection{Example: BioPortal}
-%
-%In the BRCA1 SNP example, we found ontology terms for concepts using BioPortal. Now, at the
-%time of writing BioPortal has an experimental SPARQL end point which allows us to find terms
-%in this way too. For example, to find terms related to SNPs, we can run this code in R:
-%
-%\begin{verbatim}
-%findSNPterms = paste(
-%  "PREFIX rdfs: <http://www.w3.org/2000/01/rdf-schema#>",
-%  "SELECT DISTINCT ?term ?label WHERE {",
-%  " ?term rdfs:label ?label .",
-%  "FILTER (CONTAINS (LCASE(str(?label)), \"snp\"))",
-%  "}"
-%)
-%sparql.remote(
-%  endpoint="http://sparql.bioontology.org/sparql",
-%  sparql=findSNPterms, jena=FALSE
-%)
-%\end{verbatim}
+\begin{verbatim}
+findAnnotations = paste(
+  "prefix up: <http://purl.uniprot.org/core/>",
+  "prefix ens: <http://purl.uniprot.org/ensembl/>",
+  "prefix rdfs: <http://www.w3.org/2000/01/rdf-schema#>",
+  "SELECT DISTINCT ?protein ?mnemonic ?comment WHERE {",
+  "  ?transcript up:transcribedFrom ens:ENSG00000012048 .",
+  "  ?protein a up:Protein ;",
+  "    rdfs:seeAlso ?transcript ;",
+  "    up:mnemonic ?mnemonic ;",
+  "    up:annotation ?annot .",
+  "  ?annot rdfs:comment ?comment .",
+  "}"
+)
+annotations = sparql.remote(
+  endpoint="http://beta.sparql.uniprot.org/",
+  sparql=findAnnotations, jena=FALSE
+)
+\end{verbatim}
 
@@
 
-The rrdflibs and rrdf packages bring semantic web technologies to R statistical environment.
+The rrdflibs and rrdf packages bring basic semantic web technologies to R statistical environment.
 While only a subset of Apache Jena is currently exposed, it provides key methods to deal with
-RDF data and resources.
+RDF data and resources. The examples shed some light on how it can find a place on more advanced
+analyses workflows.
 

### 2013-12-30 article.tex
Some clean up, like proper references

@@
 SIO is the Semanticscience Integrated Ontology;
-TMO is the Translational Medicine Ontology;
+TMO is the Translational Medicine Ontology~\cite{Luciano2011};
 %SO is the Sequence Types and Features Ontology;
 GO is the Gene Ontology;
-BIBO is the Bibliographic Ontology;
-CiTO is the Citation Typing Ontology;
-... Prefixes used in the term Uniform Resource Identifiers (URIs) include
+BIBO is the Bibliographic Ontology~\cite{d2009bibliographic}; and, 
+CiTO is the Citation Typing Ontology~\cite{Shotton2010}.
+
+Prefixes used in the term Uniform Resource Identifiers (URIs) include
 \textit{sio} for \url{http://semanticscience.org/resource/},
@@
 \textit{obo} for \url{http://purl.obolibrary.org/obo/},
-\textit{bibo} for \url{http://purl.org/ontology/bibo/},
-\textit{cito} for \url{http://purl.org/spar/cito/} ...
-
-We can define these prefixes in the store with the following code,
-the first prefix in the full syntax and the other namespaces
-leaving out the parameters names:
+\textit{bibo} for \url{http://purl.org/ontology/bibo/}, and
+\textit{cito} for \url{http://purl.org/spar/cito/}.
+We can add these prefixes to the store. The following code examples
+shows the full syntax in the first addition, but leaves out the
+parameters names in the following additions:
 

### 2013-12-30 article.tex
Place the examples inside the larger sections for which they are an example
Added an abstract

@@
 \begin{abstract}
+\noindent
+\textbf{Background}. Semantic Web technologies are increasingly used in biological database systems.
+The improved expressiveness show advantages in tracking provenance and allowing knowledge to be
+more explicitly annotated. The list of semantic web standards needs a complementary set of tools to
+handle data in those formats to use them in bioinformatics workflows. \\
+\textbf{Methods}. The approach proposed in this paper uses the Apache Jena library to create an
+environment where semantic web technologies can be use in the statistical environment R. The
+code is exposed as two R packages available from the Comprehensive R Archive Network (CRAN). 
+The RJava libary and a custom convenience class is used to bridge between R and the Jena library.\\
+\textbf{Results}. We here present two examples showing how the Resource Description Framework
+(RDF) and SPARQL query standards can be employed in R. The first example takes input on BRCA1 SNPs
+from a BioMart and converts this into a RDF data set. The second example runs a query on an
+experimental remote SPARQL end point provided by Uniprot, and searches textual annotations of
+proteins encoded by the BRCA1 gene. \\
+\textbf{Discussion}. The two provided library bring basic semantic web technologies to R.
+While only a subset of Apache Jena is currently exposed, it provides key methods to deal with
+RDF data and resources. The libraries are freely available from the CRAN under the 
+Affero GNU Public Licence version 3: \url{http://cran.r-project.org/web/packages/rrdf/}.
 \end{abstract}
@@
 
-\subsection{Example}
+\subsubsection{Example}
 

### 2013-12-30 article.tex
Spell checked

@@
 code is exposed as two R packages available from the Comprehensive R Archive Network (CRAN). 
-The RJava libary and a custom convenience class is used to bridge between R and the Jena library.\\
+The RJava library and a custom convenience class is used to bridge between R and the Jena library.\\
 \textbf{Results}. We here present two examples showing how the Resource Description Framework
@@
 RDF data and resources. The libraries are freely available from the CRAN under the 
-Affero GNU Public Licence version 3: \url{http://cran.r-project.org/web/packages/rrdf/}.
+Affero GNU Public License version 3: \url{http://cran.r-project.org/web/packages/rrdf/}.
 \end{abstract}
@@
 Remote SPARQLing can be used to get more information about biological entities of interest. The following
-example retreives information from the Uniprot database using the experimental SPARQL end point. In
+example retrieves information from the Uniprot database using the experimental SPARQL end point. In
 particular, it retrieves annotations for proteins encoded by the BRCA1 gene:

### 2013-12-30 article.tex
Wrote up another example, something I actually frequently do

@@
 
+\subsubsection{Example}
+
+A second example combines regular REST or -like services that return RDF with a local SPARQL query.
+For example, this is the approach used in Bioclipse to interact with many of the OpenTox
+services~\cite{Willighagen2013}, but can be used for the Open PHACTS Linked Data API
+too~\cite{Williams2012}. For example, we can get data from an OpenTox data set using
+RCurl~\cite{Hardy2010,Lang2013}, in this example data set 112 with 320 compounds,
+which originates from the ToxCast project~\cite{Dix2007}:
+
+\begin{footnotesize}
+\begin{verbatim}
+library(RCurl)
+rdfContent = getURL(
+  paste(
+    "http://apps.ideaconsult.net:8080/ambit2/dataset/112/compounds",
+    "media=text/n3",
+    sep="?"
+  ),
+  write=basicTextGatherer()
+)
+store = fromString.rdf(rdfContent, format="N3")
+compounds = sparql.rdf(store,
+  paste(
+    "PREFIX ot: <http://www.opentox.org/api/1.1#>",
+    "SELECT DISTINCT ?compound WHERE {",
+    " ?compound a ot:Compound",
+    "}"
+  )
+)
+\end{verbatim}
+\end{footnotesize}
+
+In fact, we can continue getting further detail about these compound by repeating the same
+approach on the compound URIs. Let's assume we are further interested in the SMILES string.
+This feature has the URI \url{http://apps.ideaconsult.net:8080/ambit2/feature/21753}. We
+add this as the list of features to retrieve:
+
+\begin{footnotesize}
+\begin{verbatim}
+feature = "http://apps.ideaconsult.net:8080/ambit2/feature/21753"
+getOptions = paste(
+  paste("feature_uris[]", curlEscape(feature), sep="="),
+  paste("media", curlEscape("application/rdf+xml"),sep="="),
+  sep="&"
+)
+\end{verbatim}
+\end{footnotesize}
+
+We can then use the R apply() method to retrieve the SMILES for each structure with:
+
+\begin{footnotesize}
+\begin{verbatim}
+result = apply(compounds, MARGIN=1, function(x) {
+  compound = x["compound"];
+  compoundURI = paste(compound, getOptions, sep="?")
+  print(paste("Downloading",compoundURI))
+  cmpdContent = getURL(
+    compoundURI, write=basicTextGatherer()
+  )
+  fromString.rdf(cmpdContent, format="RDF/XML", appendTo=store)
+})
+\end{verbatim}
+\end{footnotesize}
+
+Finally, we use SPARQL again to list all the SMILES strings for the compounds:
+
+\begin{footnotesize}
+\begin{verbatim}
+query =  paste(
+  "PREFIX ot: <http://www.opentox.org/api/1.1#> ",
+  "SELECT DISTINCT ?compound ?value WHERE {",
+  " ?s ?o [ ot:feature <", feature, "> ;",
+  "   ot:value ?value ",
+  " ] ;",
+  "   ot:compound ?compound . ",
+  "}",
+  sep=""    
+)
+compoundSMILESes = sparql.rdf(store, query)
+\end{verbatim}
+\end{footnotesize}
+
+This results in a matrix with compound URIs in one column and SMILES strings in another, ready for
+use by other R packages.
+
 \section{Conclusions}

### 2013-12-31 article.tex
Fixed the KI affiliation

@@
 $^1$ Department of Bioinformatics - BiGCaT, NUTRIM, Maastricht University, Maastricht, The Netherlands \\
-$^2$ Department of Molecular Toxicology, IMM, Karolinska Institutet, Stockholm, Sweden
+$^2$ Division of Molecular Toxicology, IMM, Karolinska Institutet, Stockholm, Sweden
 

### 2013-12-31 article.tex
Added ToxBank and Open PHACTS grant details

@@
 getting into contact with me.
+The research leading to these results has received funding from the European Community’s Seventh
+Framework Programme (FP7/2007–2013) under Grant Agreement n° [267042], from Cosmetics Europe, 
+and from the Innovative Medicines Initiative Joint Undertaking under grant agreement no. 115191,
+resources of which are composed of financial contribution from the European Union's Seventh Framework
+Programme (FP7/2007-2013) and EFPIA companies’ in-kind contribution.
 

### 2014-02-01 article.tex
Make the title a bit more specific: just R
The paper actually has three examples

@@
 
-\title{Accessing biological data with semantic web technologies}
+\title{Accessing biological data in R with semantic web technologies}
 \author{Egon L.\ Willighagen$^{1,2}$}
@@
 The RJava library and a custom convenience class is used to bridge between R and the Jena library.\\
-\textbf{Results}. We here present two examples showing how the Resource Description Framework
+\textbf{Results}. We here present three examples showing how the Resource Description Framework
 (RDF) and SPARQL query standards can be employed in R. The first example takes input on BRCA1 SNPs
@@
 experimental remote SPARQL end point provided by Uniprot, and searches textual annotations of
-proteins encoded by the BRCA1 gene. \\
+proteins encoded by the BRCA1 gene. The third example shows how the package can be used to handle
+RDF returned by REST-like web services of OpenTox. \\
 \textbf{Discussion}. The two provided library bring basic semantic web technologies to R.

### 2014-02-01 article.tex
A typo fix

@@
  PubMed identifier & CHEMINF & PubMed identifier & sio:CHEMINF\_000302\\
- chromosome number & TMO & chromosome nuber & tmo:TMO\_0157 \\
+ chromosome number & TMO & chromosome number & tmo:TMO\_0157 \\
  chromosomal sequence position & TMO & chromosomal sequence position & tmo:TMO\_0122 \\

### 2014-03-15 article.tex
No need to go into the REST discussion...
Added a note about general applicability

@@
 proteins encoded by the BRCA1 gene. The third example shows how the package can be used to handle
-RDF returned by REST-like web services of OpenTox. \\
+RDF returned by OpenTox web services. \\
 \textbf{Discussion}. The two provided library bring basic semantic web technologies to R.
-While only a subset of Apache Jena is currently exposed, it provides key methods to deal with
-RDF data and resources. The libraries are freely available from the CRAN under the 
+This paper only shows examples from the biology domain, but we believe the approaches are generally
+applicable. It currently only exposes a subset of key Apache Jena is exposed, but the framework
+makes it easy to make more of the library's functionality available.
+The rrdf libraries are freely available from the CRAN under the 
 Affero GNU Public License version 3: \url{http://cran.r-project.org/web/packages/rrdf/}.

### 2014-03-15 article.tex
Applied updated based on Chris' comments at https://github.com/Chris-Evelo/rrdf-paper/commit/efbbcac0fb79edd146be299a5ad5162b852737f7

@@
 \textbf{Background}. Semantic Web technologies are increasingly used in biological database systems.
-The improved expressiveness show advantages in tracking provenance and allowing knowledge to be
+The improved expressiveness shows advantages in tracking provenance and allowing knowledge to be
 more explicitly annotated. The list of semantic web standards needs a complementary set of tools to
@@
 \textbf{Methods}. The approach proposed in this paper uses the Apache Jena library to create an
-environment where semantic web technologies can be use in the statistical environment R. The
+environment where semantic web technologies can be used in the statistical environment R. The
 code is exposed as two R packages available from the Comprehensive R Archive Network (CRAN). 
@@
 This paper only shows examples from the biology domain, but we believe the approaches are generally
-applicable. It currently only exposes a subset of key Apache Jena is exposed, but the framework
-makes it easy to make more of the library's functionality available.
+applicable. It currently only exposes a subset of key Apache Jena, but the rrdf package
+makes it easy to make more of the library's functionality, such as the shortest path finding.
 The rrdf libraries are freely available from the CRAN under the 
@@
 Semantic Web technologies are finding their way into biological
-databases~\cite{Belleau2008,Chen2010,Samwald2011,Williams2012,Willighagen2013}. The recent launch of the
+databases~\cite{Belleau2008,Chen2010,Samwald2011,Williams2012,Willighagen2013}
+and the pace is increasing. The recent launch of the the
 Resource Description Framework (RDF) services at the European Bioinformatics Institute is a good example
-of the impact it has on the field, despite the adoption taking place for longer already~\cite{EBIRDFPlatform}.
+of the impact it has on the field~\cite{EBIRDFPlatform}.
 At the same time, most new databases do not yet use semantic web technologies, which may be partly
-caused by the lack of availability of tools to handle RDF data.
+caused by the lack of availability of tools to handle RDF data, and in part because of a lack
+of tutorials and illustrative examples.
 
@@
 specifications, such as the underlying RDF, RDF Schema, the Web Ontology Language, and serialization formats
-such as RDF/XML, N-TRIPLES, Notation3 (N3), and Turtle. Crucial, however, seems the specification of the SPARQL
-query language that allow extracting data from local or remote RDF triple stores~\cite{Seaborne2009}. In fact,
+such as RDF/XML, N-TRIPLES, Notation3 (N3), and Turtle. Crucial, however, seems to be the specification of the SPARQL
+query language that allows extracting data from local or remote RDF triple stores~\cite{Seaborne2009}. In fact,
 federated SPARQL queries allow extracting data from multiple triple stores at the same time.
@@
 At the same time, R has established itself as a key tool for statistics, particularly in the biological
-sciences. Many packages have been developed for this domain and are available from CRAN and Bioconductor,
+sciences. Many packages have been developed for this domain and are available from CRAN and Bioconductor~\cite{VanHage2013,Gentleman2004},
 allowing to combine various bioinformatics and data analysis approaches. These packages include,
-for example, biomaRt, which have the purpose of extracting information from a remote database.
+for example, biomaRt, which have the purpose of extracting information from a remote database~\cite{Durinck2005}.
 
@@
 Jena library and the RJava interface to Java~\cite{RMan,Jena,RJava}. Compared to the existing SPARQL
-package, this package adds general RDF data handling~\cite{VanHage2013}. Using this package we show how
+package, this package adds general RDF data handling. Using this package we show how
 data aggregation can be performed from remote SPARQL end points, and how local data in other formats
@@
 
-The rrdf packages provides RDF and SPARQL functionality by exposing that functionality from the
+The rrdf packages provide RDF and SPARQL functionality by exposing that functionality from the
 Apache Jena library~\cite{Jena}. The rrdflibs package contains the Apache Jena libraries and
-nothing more. This package is close to 7~MiB large but does not change frequently. The rrdf
-package contains the R functions that wrap around the Jena functionality and convert data structures
-where needed. This package is about 112~kiB but changes more frequently. Using this approach,
+nothing more. This package is close to 7~MiB large but is expected to change only with major new
+Jena releases. The rrdf package, however, contains the R functions that wrap around the Jena
+functionality and convert data structures where needed. This package is about 112~kiB and can
+now be updated without users having to download
+the same Jena libraries repetitively. Using this approach,
 the Comprehensive R Archive Network (CRAN) server has the least amount of download throughput.
@@
 The rrdf package provides basic and less basic functionality for aggregation and analysis of RDF data.
-The package can be install and loaded with these commands:
+The package can be installed and loaded with these commands:
 
@@
 
-To handle triples, we first need a triple store. At this moment only in-memory stores are
+To handle RDF triples, we first need a triple store. At this moment only in-memory stores are
 supported, though Jena's TDB also provides on-disk triple stores; this puts limitations to
@@
 The package supports two kinds of stores, one that has minimal ontology support, and one
-basically just handles triples. Both can be created with the \textit{new.rdf} command:
+that basically just handles triples and does not have the overhead caused by reasoning.
+Both can be created with the \textit{new.rdf} command:
 
@@
 
-When data is not read from a file or downloaded, but created from data in another format,
-such as an R matrix, individual triples can be added to a store. This works for both
-object properties and data properties. The former option links a subject to an object
+When data is not read from or downloaded downloaded as an RDF file (which is demonstrated below),
+but created from data in another format, such as an R matrix, data can be added as individual triples
+to the local store. By converting data into triples, it is later possible to make an integrated query
+that requires your local, new data and data from a remote database. Furthermore, this approach can
+also be used to expose study results as triples, such as in nanopublication format~\cite{Mons2011}.
+
+Triples can be created for both object and data properties. The former option links a subject to an object
 resource, while the latter links a subject to a literal value:
@@
 
-Data can be loaded from file by providing a file name and a format ("RDF/XML", "TURTLE",
+Additionally, data can be loaded from file by providing a file name and a format ("RDF/XML", "TURTLE",
 "N-TRIPLES", and "N3"):
@@
 store = load.rdf("file.n3", format="N3")
+\end{verbatim}
+\end{footnotesize}
+
+Similarly, triples in a local store can also be stored as an RDF file, for example with:
+
+\begin{footnotesize}
+\begin{verbatim}
+store = save.rdf(store, filename="copy.n3", format="N3")
 \end{verbatim}

### 2014-03-15 article.tex
Fixed a random insert error

@@
 library(biomaRt)
-mart = biomaNoy2009Rt::useMart(biomart="snp", dataset="hsapiens_snp")
+mart = biomaRt::useMart(biomart="snp", dataset="hsapiens_snp")
 brca1 = c("rs16940","rs16941", "rs16942", "rs799916", "rs799917")

### 2015-02-18 article.tex
Updated for today's rrdf and rrdflibs versions

@@
 makes it easy to make more of the library's functionality, such as the shortest path finding.
-The rrdf libraries are freely available from the CRAN under the 
-Affero GNU Public License version 3: \url{http://cran.r-project.org/web/packages/rrdf/}.
+The rrdf packages are freely available from GitHub under the 
+Affero GNU Public License version 3: \url{https://github.com/egonw/rrdf/}.
 \end{abstract}
@@
 Apache Jena library~\cite{Jena}. The rrdflibs package contains the Apache Jena libraries and
-nothing more. This package is close to 7~MiB large but is expected to change only with major new
+nothing more. This package is about 7~MiB large but is expected to change only with major new
 Jena releases. The rrdf package, however, contains the R functions that wrap around the Jena
@@
 now be updated without users having to download
-the same Jena libraries repetitively. Using this approach,
-the Comprehensive R Archive Network (CRAN) server has the least amount of download throughput.
+the same Jena libraries repetitively. Using this approach, the download size is minimized.
 The Jena functionality is accessible using the rJava library~\cite{RJava} which handles loading
@@
 
-The source code of the package is available from GitHub at \url{http://github.com/egonw/rrdf/} and
-binary packages are available from CRAN at \url{http://cran.r-project.org/web/packages/rrdf/}.
-The first patch is from March 2011 while the most recent patches are from late 2013 when the
-rrdflibs package was updated for Apache Jena 2.11.
+The source code of the package is available from GitHub at \url{http://github.com/egonw/rrdf/}.
+Binary packages are available here too and can be installed easily from GitHub using the 
+devtools package~\cite{devtools}. The first patch of the library is from March 2011 while the
+most recent patches are from late 2013 when the rrdflibs package was updated for Apache Jena 2.11.
+Early 2015 the rrdflibs package was updated to Apache Jena 2.12.1.
 
